Confronting preferential sampling in wildlife surveys: diagnosis and model-based triage \footnotemark[2]

### 2016-11-18 manuscript/ReviewerResponses.tex

\documentclass[12pt,fleqn]{article}
%\documentclass[12pt,a4paper]{article}
\usepackage{amsmath,epsfig,epsf,psfrag,color,ecology,natbib}
%\usepackage{a4wide,amsmath,epsfig,epsf,psfrag}
\usepackage[nolists]{endfloat}
\bibliographystyle{ecology}

%  put your commands here

\def\be{{\ensuremath\mathbf{e}}}
\def\bx{{\ensuremath\mathbf{x}}}
\def\bftheta{{\ensuremath\boldsymbol{\theta}}}
\def\bfeta{{\ensuremath\boldsymbol{\eta}}}
\def\bfdelta{{\ensuremath\boldsymbol{\delta}}}
\def\bfnu{{\ensuremath\boldsymbol{\nu}}}
\def\bfmu{{\ensuremath\boldsymbol{\mu}}}
\def\bfSigma{{\ensuremath\boldsymbol{\Sigma}}}
\def\bfB{\textbf{B}}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}

@@

\bf \hspace{-.32in} Conn et al. responses to reviewer comments

%  make sure that the document has 25 lines per page (it is 12 pt)

\setlength{\textheight}{575pt} \setlength{\baselineskip}{23pt}

%  IMPORTANT -- place the \title, \author, etc, \begin{summary}...
%  \end{summary} \keywords statements in the order demonstrated below
\rm
Here, we provide line by line responses to associate editor and reviewer comments, suggestions, and edits.  For completeness, we include all reviewer comments in addition to our responses (our responses are in blue).  In addition, our revision includes copy editing corrections from our editorial office and some clarifications in response to additional comments we received from an interested reader. We provide a marked up PDF version of our manuscript showing all changes made.

@@

[From the comments to author section]
Your paper definitely focuses on an important issue, and provides ways to address the problem. Besides the issues raised by the referee, I think you should relate your paper a bit more to the ecological literature, in particular on SDMs (see Edwards et al. 2006). I think in particular that in addition to the types of sampling you discuss, one could also consider sampling designs that are optimal to estimate response curve (eg Albert et al. 2010 - sorry for the self-citation...). Ecological studies may have different objectives, and I think you should try to make your paper relevant to as many readers as possible.

Albert, C. H., N. G. Yoccoz, T. C. Edwards, C. H. Graham, N. E. Zimmermann, and W. Thuiller. 2010. Sampling in ecology and evolution – bridging the gap between theory and practice. Ecography 33:1028-1037.
Edwards, T. C., D. R. Cutler, N. E. Zimmermann, L. Geiser, and G. G. Moisen. 2006. Effects of sample survey design on the accuracy of classification tree models in species distribution models. Ecological Modelling 199:132-141.

\textcolor{blue}{We thank the editor for pointing us to these articles - our knowledge of the species distribution modeling literature is admittedly a little lacking.  These papers are certainly germane to the present study, and we have included additional material in the introduction (2nd paragraph) that talks about (1) implications of design vs. model-based inference, and (2) the importance of the particular objective function (e.g., minimizing variance, estimating species-habitat relationships) for optimal design.}

Additionally, we have made the following changes to try to make the manuscript accessible to a wider audience:
\begin{itemize}
  \item 
\end{itemize}

@@

I have difficulty following your notation at several points throughout the manuscript, which are outlined
below.

On line 41, you state $Z_i$ is the state variable of interest (e.g., occupancy or abundance), which I take will
be the notation you use throughout given the section header. Your specication for a count model on eqn
4 therefore becomes confusing. In particular, if $Z_i$ represents abundance, for reasonable values of $A_i$ and $p_i$ you could have random realizations $Y_i$ from a Poisson distribution that are greater than abundance, $Z_i$. If $Z_i$ does not represent abundance, what state process is represented here?

\textcolor{blue}{
  We have replaced the Poisson observation model with a binomial one in eqn 4 to make this clearer, and have indicated dependence on $Z_i$ through conditioning.  The reason that we had the Poisson in there originally is that the models we fit to data use the fact that the marginal distribution for $Y_i$ given the Poisson-Binomial mixture is itself a Poisson distribution.  However, we think it will be clearer to present the model hierarchically at this stage and then to describe our marginalization (something many ecologists may not care about) when describing software.  We have created a new ``Software" methods subsection which describes this in greater detail. Importantly, we describe how predictions are made; when detection probability is equal to 1.0, predictions are made using the formulation $\hat{N}_i = Y_i + (1-A_i)\exp(\hat{\mu}_i)$.  This framework incorporates a finite population correction and is similar to that employed by \citet{JohnsonEtAl2010} in the context of point process likelihood analysis of distance sampling data.  It also prevents predicted abundance in a cell from being less than the observed count.
}

On line 61, you indicate $\bfmu$ and $\bfnu$ represent log-scale abundance (associated with {\bf Z}) and the logit of the probability of sampling (associated with {\bf R}). For equation 5, however, you use $\bfmu_i$ to represent the mean value of a Gaussian process, which created some confusion. You might consider more generally allowing $\bfmu$ to be a function of covariates influencing the state variable of interest, and $\bfnu$ to be a function of covariates
influencing placement of sampling locations. You could then replace $\psi_i$ in equation 6 with $\nu_i$, which would make it immediately clear you were discussing factors influencing the probability of site placement.

\textcolor{blue}{
  We appreciate the reviewer's attempt to help us streamline notation and to make things clearer for the reader.  We have implemented the basic premise of this idea, although some of the particulars differ from the reviewer's suggestion.  Specifically we now describe $\bfmu$ and $\bfnu$ as indicating ``the link-scale process of interest and the logit of the probability of sampling, respectively.  As with $\bfmu$, this would mean $[Z_i] = f(\exp(\mu_i))$, whereas the distribution of sampling locations can be represented as $[R_i] = f(\text{logit}^{-1}(\nu_i))$, where $\mu_i$ and $\nu_i$ can be written as functions of regression coefficients, habitat covariates, and spatially autocorrelated random effects."  Then in the next section we use $\mu_i$ in eq. 5 and $\nu_i$ in eq. 6, although we retain $xi_i$, since in the next section we also wish to write $\nu_i$ in terms of shared spatial effects.  This is all a little tricky, in that the \citet{DiggleEtAl2010} paper considers a continuous support model (whereas ours is discrete), and the structure is slightly different.  Admittedly, maintaining recognizable notation while keeping precise definitions is a challenge.
}

I understand the model you are specifying under the ``simulation study" subheading, but the notation is
different from that presented in the rest of the paper (i.e., use of ``=" vs. ``$\sim$"). I have no preference for
the notation used, but it seems like consistent notation would be more reader friendly. The same comment
holds true for the model presented under the subheading ``bearded seal count surveys".

\textcolor{blue}{
  We agree.  We've changed all distributions to brackets and have eliminated the ``$\sim$" notation.
}

Finally, your definition of $Z_i$ in the ``bearded seal count surveys" section seems inconsistent with the general
definition of $Z_i$ at the beginning of your manuscript. I think of state variables as presence / absence,
abundance, survival, etc. This definition for $Z_i$ seems more closely related to the expected value of a state
process.

\textcolor{blue}{
  We've rewritten the model here in terms of the Poisson-binomial hierarchical mixture.  In the following ``Software" section, we then indicate that we marginalize over $\bfZ$ when performing ML inference.
}

{\bf Other Comments}

Line 65: This sentence seems to discount the value of good study design, which is still important for model-
based estimation.

\textcolor{blue}{
We've altered text to be a little less cavalier, now stating that 
``Design-based estimation requires that surveys rely on a pre-planned survey design selected probabilistically from an underlying sampling frame \citep{Cochran1977}.  Although randomization can increase legitimacy, model-based estimation is not formally subject to this requirement."
}

Line 77: It's unclear to me why the distinction between fine and coarse-scale preferential sampling is important. They both seem to represent the process of dependence between sampling locations and the process
of interest.

\textcolor{blue}{
As the size of the sampling unit decreases (with the limit being a point process model in continuous space) there is indeed little difference.  But practically, it seems likely that one could use a reasonable design to select sampling units, but still open themselves to bias if they target non-representative habitat.  However, this point is well taken, and we don't think the distinction necessarily needs to prominence that we've given it.  We now address this topic for the first time in the discussion.
}

Line 86: I do not see how the definition offered here differs from your initial description of preferential
sampling, unless the environmental covariate associated with the probability a location is selected is itself
independent from the state variable of interest, which seems unlikely in general.

\textcolor{blue}{
This is an extremely important point that we may have failed to get across.  Selecting sampling units based on a covariate related to abundance is just fine, \textit{provided} that the covariate is also used in the model for animal density/occurrence. In essence, we really only have preferential sampling if there the probability of sampling is related to the residuals (say $\hat{N_i} - N$) from a fitted SDM.  We've produced an additional figure (now fig. 1) that attempts to provide intuition about the distinction between sampling that is covariate-dependent and sampling that is preferential.
}

Line 186: I understand why you fit a relatively simple abundance model in this case. However, given the
difficulties you encountered in incorporating even modestly complex structure on B, how well do you suppose
your model will perform when you attempt to correct for nuisance processes?

\textcolor{blue}{
We don't know the answer to this question, although it seems like an important one.  In the discussion, we now suggest this as a topic for future research.
}

Line 195: You only describe 4 models below, not 6.

\textcolor{blue}{Changed}

Simulation study results: can you report bias for other parameters (e.g. slope parameters) in your model,
even if this is relegated to an appendix?

\textcolor{blue}{Yes, we have created an additional appendix table (S3) that summarizes bias in $b$ (the preferential sampling parameter), as well as bias in the species-habitat parameters.  We don't look at bias in the sampling probability parameters because data were simulated by sampling without replacement (with a fixed number of sampled survey units), whereas the fitted model was Bernoulli (which assumes the number of sampled units is random).  Thus parameters would not be directly comparable.  We also provide convergence statistics in this appendix. %We note that the bias in the slope parameter of the species-habitat relationship %is likely because of the well-known confounding that occurs between regression %coefficients and spatially autocorrelated random effects in spatial models %\citep[e.g.][]{ReichEtAl2006,HodgesReich2010}.
}

Paragraph starting on line 219: I find this paragraph very confusing. In particular, it is unclear to me exactly
what you mean when you state you ``...refitted $M_{cov=1;b=0}$ to produce an estimate of apparent abundance
without this feature. Specifically, we refitted the model with 20 pseudo-absences in this portion of the study
area to better inform abundance-covariate relationships". Does Fig. 5 reflect this re-fitted model? Why not
fit all models in this manner for consistency?

\textcolor{blue}{We now include these known zero values in all models.}

Line 225: a model-averaged estimate of abundance?

\textcolor{blue}{After correcting the link function in present model runs, it no longer makes sense to present a model-averaged estimate.}

Table 1: I find these tables are easier to read if models are listed in decreasing order of $\Delta$AIC.

\textcolor{blue}{We now present models in order of $\Delta$AIC.}

Fig. 3: In my version, the blue and green dashed lines do not show up in the legend.

\textcolor{blue}{Strange, this is not the case in our version.  Guessing it is an issue with PDF compilation on the MEE submission website.  We'll work with copy editors to ensure it looks right as it heads into production.}

### 2016-11-18 manuscript/pref_sampling_revision.tex
Commit during ms. revisions and change to link function; simulations now running with some error trapping and bounds set on parameter space.  Still getting some negative estimates using sdreport with bias.correct=TRUE when b is estimated.

\address{\affilnum{1}National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA National Marine Fisheries Service, 7600 Sand Point Way NE, Seattle, WA 98115 USA; \affilnum{2} Fisheries Resource Assessment and Monitoring Division (FRAM), Northwest Fisheries Science Center, National Marine Fisheries Service (NMFS),
NOAA, 2725 Montlake Boulevard E, Seattle, WA 98112, USA}

@@

\item We develop a hierarchical statistical modeling framework for detecting and alleviating the biasing effects of preferential sampling in species distribution models fitted to count data.  The approach works by jointly modeling wildlife state variables and the locations selected for sampling, and specifying a dependent correlation structure between the two models.

\item Using simulation, we show that moderate levels of preferential sampling can lead to large (e.g. $40\%$) bias in estimates of animal density, and that our modeling approach can considerably reduce this bias.

@@

\def\VAR{{\rm Var}\,}
\def\COV{{\rm Cov}\,}
\def\Prob{{\rm P}\,}
\def\bfX{\bf X}
\def\bfbeta{\boldsymbol{\beta}}
\def\bfdelta{\boldsymbol{\delta}}
\def\bfeta{\boldsymbol{\eta}}

@@

One of the main advantages of using SDMs is that one is no longer beholden to predetermined sampling frames, and can potentially use data gathered from non-randomized designs or platforms of opportunity to make inferences about animal populations \citep{JohnsonEtAl2010}.  However,
in a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling disproportionately targets locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.  In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance. For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.  Bias might also arise if surveying effort is higher near bases of operations, and if animal abundance is higher (or lower) near bases of operations than elsewhere in the landscape.

In this article, we explore potential for bias in SDMs resulting from preferential sampling (hereafter, PS), and describe several model-based approaches for detecting and correcting for such biases.  We start by describing a common currency for notation and basic model structures considered in this paper.  Second, we review PS bias in a mathematical light, and describe prior approaches to coping with its effects. Third, we introduce a novel generalization of previously proposed PS models, allowing the investigator to jointly model animal encounter data and the locations chosen for sampling, including possible dependence structure between these two types of observations.  Fourth, we conduct a simulation study to examine the performance of traditional SDMs and our newly developed PS model when data are gathered preferentially.  Finally, we demonstrate our modeling approach by analyzing aerial survey counts of bearded seals (\textit{Erignathus barbatus}) in the Bering Sea.

@@

We focus here exclusively on discrete space (areal) models for animal encounter data as these seem to be the dominant form used in design and analysis of animal population surveys, although we note that PS is likely to affect analyses similarly regardless of the choice of spatial domain.  We suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$). Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if location $i$ is sampled (i.e. if $U_i \in \mathcal{S}$), and is 0 otherwise.  To formulate a ``traditional" SDM, one could then write animal abundance or occurrence as a stochastic realization of a probability mass function $f()$:
\begin{eqnarray}
  \label{eq:process}
  Z_i & \sim & f(g^{-1}(\mu_i)).
\end{eqnarray}
In this example, $Z_i$ denotes the state variable of interest (e.g. occupancy or abundance), $g()$ is a link function (e.g. probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value.  In applications described in this paper, we write the intensity as
\begin{eqnarray}
\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i,
 \label{eq:mu}
\end{eqnarray}
where $\beta_0$ is an intercept parameter, ${\bf x}_i$ is a row vector of $m$ predictive covariates associated with site $i$, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_m \}$ is a column vector of $m$ regression parameters, and $\boldsymbol{\delta}=\{ \delta_1,\delta_2,\hdots,\delta_S \}$ are spatially autocorrelated random effects.  For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\delta_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g. conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{Reich2006,Hughes2013}.

The model for $Z_i$ describes variation in the process of interest and is often described as the ``process'' model.  However, it is usually impossible to observe the system perfectly even in locations where sampling occurs, so it is customary to include an observation model describing incomplete detection.  For occupancy studies, the response variable $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}:
\begin{eqnarray}
  Y_i & \sim & \text{Bernoulli}(Z_i p_i).
\end{eqnarray}
 Here, the detection probability $p_i$ is possibly a function of survey and observer specific covariates. Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$.   For count surveys, a possible model is
\begin{eqnarray}
  \label{eq:obs_pois}
  Y_i & \sim & \text{Poisson}(Z_i A_i p_i),
\end{eqnarray}
where the $Y_i$ now represents the count of animals obtained while surveying unit $i$, $A_i$ denotes the proportion of sample unit $i$ that is surveyed, and $p_i$ gives detection probability.  Additional information will often be needed to estimate $p_i$ in this context, such as data from double observers, distance observations, or double sampling \citep[see e.g.][]{BucklandEtAl2001,Royle2004,Borchers2006,ConnEtAl2014}.

For the remainder of this treatment, we use bold symbols to denote vector-valued quantities or matrices.  We also use standard bracket notation to denote probability mass and density functions.  For instance $[{\bf Z}]$ denotes the marginal probability mass function for ${\bf Z}$, and  $[{\bf Z}|{\bf Y}]$ represents the conditional distribution of ${\bf Z}$ given ${\bf Y}$.  We use $\boldsymbol{\mu}$ and $\boldsymbol{\nu}$ to denote log-scale abundance and the logit of the probability of sampling, so that $Z_i \sim f(\mu_i)$, and $R_i \sim f(\nu_i)$.  We use the notation $Z_i$ when describing the state process in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.

@@

One of the appealing aspects of model-based estimation is that there is no requirement that surveys rely on a pre-planned survey design selected probabilistically from an underlying sampling frame.  For instance, investigators can reallocate sampling effort if weather or logistics preclude surveying in a desired location.  This can be a crucial advantage in surveys covering large areas with frequent inclement weather.  It also opens the door for using platforms of opportunity, presence only, and citizen science data for estimation.

However, the manner in which effort is ultimately allocated can potentially have profound influence on SDM estimator performance.  With respect to nonrandom sampling, two possible problems seem particularly likely in discrete spatial domains: coarse scale preferential sampling (CSPS), and fine scale preferential sampling (FSPS) (Fig. \ref{fig:pref}).  FSPS arises when the observations taken at a  particular sampling unit are non-random with respect to the density of animals within that sampling unit.  For instance, when allocating line transect survey effort, it may be tempting to place the transect in a manner that targets habitat or landscape features that maximize the number of animals that will be encountered.  However, this strategy will clearly lead to positive bias when estimating density or abundance.

By contrast, CSPS (hereafter, PS), the primary focus of this article, arises when the locations being sampled and the process of interest (e.g. density, occupancy) are conditionally dependent given modeled covariates \citep{DiggleEtAl2010}.  For instance, PS can occur when the investigator uses a priori knowledge or observations of the state variable obtained during sampling to allocate survey effort in places where abundance or occurrence is known to be high. \citet{DiggleEtAl2010} showed that this type of PS can lead to bias when this extra information is not included in models for the state variable of interest.  Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$, where \textbf{R} is an indicator vector whose elements $R_i$ are 1.0 if sampling unit $i$ is sampled and are zero otherwise.  We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.  For instance, \citet{MerckxEtAl2011} use the term ``preferential sampling" to refer to the process of visiting some sites more often than others, while \citet{ManceurKuhn2014} define it as occurring when the locations selected for sampling are a function of an environmental covariate.  Neither of these latter conditions are problematic outside of the specialized field of presence-only modelling.

@@

As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS.  However, in many ecological applications, predictive covariates explain only a small portion of variation present in the data.  If the locations selected for sampling are a function of some unmodelled factor related to abundance (intentionally or unintentionally) , bias may still occur.  Despite the clear potential for bias in SDMs, there are few examples where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.  One exception is
\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs.  However, they did not attempt to account for PS in their models.

@@

Several authors have attempted model-based corrections for PS in the statistical literature. For Gaussian models in a continuous spatial domain, \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} jointly modeled the locations that are chosen for sampling and the underlying random field of interest.  In particular, they expressed sampled locations as an inhomogeneous Poisson point process where the underlying log-scale intensity depended linearly on spatially-referenced random field values.  For instance, writing observations of the spatial random field at a location $i$ as
\begin{eqnarray}
   Z_i & = & \mu_i + \epsilon_i,
   \label{eq:Z_cont}
\end{eqnarray}
the spatially continuous relative intensity ($\psi_i$) of sampling locations at $i$ could be written as
\begin{eqnarray}
   \psi_i & \propto & \exp(\xi_i + b \mu_i).
   \label{eq:p_cont}
\end{eqnarray}
Here, the parameter $b$ describes the level of PS; $b=0$ implies no PS, $b>0$ implies a greater level of sampling in locations where the spatial process (e.g. animal density) is high, and $b<0$ implies greater sampling where the spatial process is low.  Importantly, when explanatory covariates are used in models for $\mu_i$ and $\xi_i$, \citet{PatiEtAl2011} show that ``. . . accounting for informative sampling is only necessary when there
is an association between the spatial surface of interest and the sampling density that cannot be
explained by the shared spatial covariates."  \citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, where the log of a nonparameteric estimate of sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in Eq. \ref{eq:Z_cont}, finding that this approach helped reduce bias associated with PS, but did not perform as well as the full joint model.

>>>>

\address{\affilnum{1}Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA, National Marine Fisheries Service, 7600 Sand Point Way NE, Seattle, WA 98115 USA; \affilnum{2} Fisheries Resource Assessment and Monitoring Division (FRAM), Northwest Fisheries Science Center, NOAA, National Marine Fisheries Service,
2725 Montlake Boulevard E, Seattle, WA 98112, USA}

@@

\item We develop a hierarchical statistical modeling framework for detecting and alleviating the biasing effects of preferential sampling in species distribution models fitted to count data.  The approach works by jointly modeling wildlife state variables and the locations selected for sampling and specifying a dependent correlation structure between the two models.

\item Using simulation, we show that even moderate levels of preferential sampling can lead to large (e.g. $40\%$) bias in estimates of animal density and that our modeling approach can considerably reduce this bias.

@@

\def\VAR{{\rm Var}\,}
\def\COV{{\rm Cov}\,}
\def\Prob{{\rm P}\,}
\def\bfX{\bf X}
\def\bfbeta{\boldsymbol{\beta}}
\def\bfdelta{\boldsymbol{\delta}}
\def\bfeta{\boldsymbol{\eta}}
\def\bfnu{\boldsymbol{\nu}}
\def\bfmu{\boldsymbol{\mu}}

@@

One of the main advantages of using SDMs is that one is less beholden to predetermined sampling frames and can potentially use data from non-randomized designs or opportunistic sampling to make inferences about animal populations \citep{JohnsonEtAl2010}. Several authors have shown that non-randomized designs can have poorer predictive ability than randomized designs when used to estimate species distributions \citep[e.g.][]{EdwardsEtAl2006,AlbertEtAl2010}, but that the optimal design may depend on the question of interest.  For instance, the optimal design may differ depending on whether the inferential goal is to maximize predictive ability or to estimate the relationship between species presence and an important habitat covariate \citep{AlbertEtAl2010}.

Less precise predictions may be a sacrifice one is willing to make if it is the only cost incurred when using data at non-random locations to make inferences about species distributions.  After all, such data are often all that are available for certain taxa and locations (e.g. from citizen science programs), and are certainly less expensive to acquire than data from randomized, scientific surveys.  However, recent research suggests that one may be opening oneself up to bias as well, which is potentially much more concerning than loss of precision. In a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling disproportionately targets locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.  In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance. For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.  Bias might also arise if survey effort is higher near bases of operations or features such as roads, particularly if animal abundance is higher (or lower) near these locations than elsewhere in the landscape.  In our view, this phenomenon, termed ``preferential sampling," is potentially much more problematic, because estimates of animal abundance and occurrence from SDMs could be severely overstated.

In this article, we explore potential for bias in SDMs resulting from preferential sampling (hereafter, PS) and describe several model-based approaches for detecting and correcting for such biases.  We start by describing a common currency for notation and basic model structures considered in this paper.  Second, we review PS bias in a mathematical light, and describe prior approaches to coping with its effects. Third, we introduce a novel generalization of previously proposed PS models, allowing the investigator to jointly model animal encounter data and the locations chosen for sampling, including possible dependence structure between these two types of observations.  Fourth, we conduct a simulation study to examine the performance of traditional SDMs and our newly developed PS model when data are gathered preferentially.  Finally, we demonstrate our modeling approach by analyzing aerial survey counts of bearded seals (\textit{Erignathus barbatus}) in the Bering Sea.

@@

We focus here exclusively on discrete space (areal) models for animal encounter data as these seem to be the dominant form used in design and analysis of animal population surveys, although we note that PS is likely to affect analyses similarly regardless of the choice of spatial domain.  We suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$). Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if location $i$ is sampled (i.e. if $U_i \in \mathcal{S}$), and is 0 otherwise.  To formulate a ``traditional" SDM, one could then write animal abundance or occurrence as a stochastic realization of a probability mass function $f()$:
\begin{eqnarray}
  \label{eq:process}
  [Z_i] & = & f(g^{-1}(\mu_i)).
\end{eqnarray}
In this example, $Z_i$ denotes the state variable of interest (e.g. occupancy or abundance), $g()$ is a link function (e.g. probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value which can itself be written as a function of habitat covariates, regression coefficients, or spatially autocorrelated random effects.  We use the bracket notation (e.g. $[Z_i]$) to denote a probability mass or probability density function (in the case of discrete or continuous distribution, respectively). In applications described in this paper, we write the intensity as
\begin{eqnarray}
\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i,
 \label{eq:mu}
\end{eqnarray}
where $\beta_0$ is an intercept parameter, ${\bf x}_i$ is a row vector of $m$ predictive covariates associated with site $i$, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_m \}$ is a column vector of $m$ regression parameters, and $\boldsymbol{\delta}=\{ \delta_1,\delta_2,\hdots,\delta_S \}$ are spatially autocorrelated random effects.  For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\delta_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g. conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{Reich2006,Hughes2013}.

The model for $Z_i$ describes variation in the process of interest and is often described as the ``process'' model.  However, it is usually impossible to observe the system perfectly even in locations where sampling occurs, so it is customary to include an observation model describing incomplete detection.  For occupancy studies, the response variable $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}:
\begin{eqnarray}
  [Y_i|Z_i] & = & \text{Bernoulli}(Z_i p_i).
\end{eqnarray}
 Here, the detection probability $p_i$ is possibly a function of survey and observer specific covariates. Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$.   For count surveys, a possible model is
\begin{eqnarray}
  \label{eq:obs_binomial}
  [Y_i|Z_i] & = & \text{Binomial}(Z_i, A_i p_i),
\end{eqnarray}
where the $Y_i$ now represents the count of animals obtained while surveying unit $i$, $A_i$ denotes the proportion of sample unit $i$ that is surveyed, and $p_i$ gives detection probability.  Additional information will often be needed to estimate $p_i$ in this context, such as data from double observers, distance observations, or double sampling \citep[see e.g.][]{BucklandEtAl2001,Royle2004,Borchers2006,ConnEtAl2014}.

For the remainder of this treatment, we use bold symbols to denote vector-valued quantities or matrices.  We also use ``$|$" to specify that a distribution is conditional; for instance, if $[{\bf Y}]$ denotes the marginal probability mass function for ${\bf Y}$, then  $[{\bf Y}|{\bf Z}]$ represents the conditional distribution of ${\bf Y}$ given ${\bf Z}$.  We use $\bfmu$ and $\bfnu$ to indicate the link-scale process of interest and the logit of the probability of sampling, respectively.  As with $\bfmu$, this would mean $[Z_i] = f(\exp(\mu_i))$, whereas the distribution of sampling locations can be represented as $[R_i] = f(\text{logit}^{-1}(\nu_i))$, where $\mu_i$ and $\nu_i$ can be written as functions of regression coefficients, habitat covariates, and spatially autocorrelated random effects.  We use the notation $Z_i$ when describing the state process in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.

@@

Design-based estimation requires that surveys rely on a pre-planned survey design selected probabilistically from an underlying sampling frame \citep{Cochran1977}.  Although randomization can increase legitimacy, model-based estimation is not formally subject to this requirement.  For instance, investigators can reallocate sampling effort if weather or logistics preclude surveying in a desired location and still generate model-based estimates of abundance or occupancy.  This can be a crucial advantage in surveys covering large areas with frequent inclement weather.  It also opens the door for using presence only, citizen science, or opportunistically collected data for estimation.

However, the manner in which effort is allocated can potentially have profound influence on SDM estimator performance.   Preferential sampling (PS) arises when the locations being sampled and the process of interest (e.g. density, occupancy) are conditionally dependent given modeled covariates \citep{DiggleEtAl2010}.  For instance, PS can occur when the investigator tends to sample more in places where a species has already been observed, or is known to be abundant. \citet{DiggleEtAl2010} showed that this type of PS can lead to bias when this extra information is not included in models for the state variable of interest.  Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$, where \textbf{R} is an indicator vector whose elements $R_i$ are 1.0 if sampling unit $i$ is sampled and are zero otherwise (see e.g. fig. \ref{fig:PS_example}).  We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.  For instance, \citet{MerckxEtAl2011} use the term ``preferential sampling" to refer to the process of visiting some sites more often than others, while \citet{ManceurKuhn2014} define it as occurring when the locations selected for sampling are a function of an environmental covariate.  Neither of these latter conditions are problematic outside of the specialized field of presence-only modelling.

@@

As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS (e.g. fig. \ref{fig:PS_example}).  However, in many ecological applications, predictive covariates explain only a small portion of variation present in the data.  If the locations selected for sampling are a function of some unmodelled factor related to abundance (intentionally or unintentionally) , bias may still occur.  Despite the clear potential for bias in SDMs, there are few examples where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.  One exception is
\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs.  However, they did not attempt to account for PS in their models.

@@

Several authors have attempted model-based corrections for PS in the statistical literature. For Gaussian models in a continuous spatial domain, \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} jointly modeled the locations that are chosen for sampling and the underlying random field of interest.  In particular, they expressed sampled locations as an inhomogeneous Poisson point process where the underlying log-scale intensity depended linearly on spatially-referenced random field values.  For instance, writing observations of the spatial random field at a location $i$ as
\begin{eqnarray}
   Z_i & = & \mu_i + \epsilon_i,
   \label{eq:Z_cont}
\end{eqnarray}
the log-scale relative intensity of sampling locations at $i$ ($\nu_i$) could be written as
\begin{eqnarray}
   \nu_i & \propto & \xi_i + b \mu_i,
   \label{eq:p_cont}
\end{eqnarray}
where $\mu_i$ and $\xi_i$ includes explanatory covariates.
Here, the parameter $b$ describes the level of PS; $b=0$ implies no PS, $b>0$ implies a greater level of sampling in locations where the spatial process (e.g. animal density) is high, and $b<0$ implies greater sampling where the spatial process is low.  Importantly, when explanatory covariates are used in models for $\mu_i$ and $\xi_i$, \citet{PatiEtAl2011} show that ``. . . accounting for informative sampling is only necessary when there
is an association between the spatial surface of interest and the sampling density that cannot be
explained by the shared spatial covariates" (see e.g. fig. \ref{fig:PS_example}).    \citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, where the log of a nonparameteric estimate of sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in eqn. (\ref{eq:Z_cont}), finding that this approach helped reduce bias associated with PS, but did not perform as well as the full joint model.
